---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r  echo=TRUE}
activity <- read.csv("activity.csv")

#ignore NA values
activityC <- activity[complete.cases(activity),]
```
## What is mean total number of steps taken per day?

```{r echo=TRUE}
#calculate the number of steps on each day
dailySteps <- aggregate(activityC$steps, by=list(activityC$date),FUN=sum)

#specifiy the column names
colnames(dailySteps) <- c("Date","steps")

#plot the histogram
hist(dailySteps$steps,breaks=53,plot=TRUE,xlab="Daily Steps", main=paste("Histogram of number of steps on each day"))

#calculate mean of number of steps taken each day
dailyStepsMean <- aggregate(activityC$steps, by=list(activityC$date),FUN=mean, na.action = na.omit)
colnames(dailyStepsMean) <- c("Date","Mean")
dailyStepsMean$Mean = round(dailyStepsMean$Mean)

#calculate mean of number of steps taken each day
dailyStepsMedian <- aggregate(activityC$steps, by=list(activityC$date), FUN=median)
colnames(dailyStepsMedian) <- c("Date","Median")
dailyStepsMedian = round(dailyStepsMedian$Median)

#Merge mean and median in one table
MeanAndMedianSteps <- merge(dailyStepsMean,dailyStepsMedian)

#report mean and median
print(MeanAndMedianSteps, type="html")
```

## What is the average daily activity pattern?
```{r echo=TRUE}
#calculate average of number of steps taken by time interval (accross all days)
dailyStepsIntervalMean <- aggregate(activityC$steps, by=list(activityC$interval),FUN=mean, na.action = na.omit)

colnames(dailyStepsIntervalMean) <- c("TimeInterval","Steps")

plot(dailyStepsIntervalMean$TimeInterval,dailyStepsIntervalMean$Steps, main="Time series of the 5-minute interval and average number of steps taken", xlab = "Time Interval", ylab="Steps taken (averaged across all days)", type="l")
 
```
The above time-series plot shows that maximum number of steps are taken in the morning around 0800 to 0900 hours 

## Imputing missing values
```{r}
#calculate the number of rows in the original dataset
RowswithNa <- nrow(activity)
RowswithoutNa <- nrow(activityC)  # activityC is a dataset with complete cases.

#calculate the difference in rows
DiffinRows <- RowswithNa - RowswithoutNa
```

The total number of missing values in the dataset (i.e. the total number of rows with NAs are `r DiffinRows`.

```{r}
#make a copy of the original dataset
ActivityFillNA <- activity

#for each interval that shows NA as number of steps taken, fill it with average steps taken at that interval (obviously rounded!)
for (val in 1:nrow(ActivityFillNA))
{
    if(is.na(ActivityFillNA[val,1]))
    {
        stepsValue = subset(dailyStepsIntervalMean, TimeInterval == ActivityFillNA[val,3])
        ActivityFillNA[val,1] = round(stepsValue$Steps,digits=0)
    }
}

```
## plot the histogram using imputed data set

```{r}

#calculate the number of steps on each day using new dataset
dailyStepsNew <- aggregate(ActivityFillNA$steps, by=list(ActivityFillNA$date),FUN=sum)

#specifiy the column names
colnames(dailyStepsNew) <- c("Date","steps")

#plot the histogram
hist(dailyStepsNew$steps,breaks=53,plot=TRUE,xlab="Daily Steps", main=paste("Histogram of number of steps on each day"))


#calculate average of number of steps taken by time interval (accross all days)
dailyStepsIntervalMeanIm <- aggregate(ActivityFillNA$steps, by=list(ActivityFillNA$interval),FUN=mean)

colnames(dailyStepsIntervalMeanIm) <- c("TimeInterval","Steps")

#plot the time series across all days using imputed data set
plot(dailyStepsIntervalMeanIm$TimeInterval,dailyStepsIntervalMeanIm$Steps, main="Time series using imputed data set", xlab = "Time Interval", ylab="Steps taken (averaged across all days)", type="l")

```

## Are there differences in activity patterns between weekdays and weekends?

```{r}
#add weekday or weekend days factor to the data set
ActivityFillNA$wdays <- as.factor(ifelse(weekdays(as.Date(ActivityFillNA$date)) %in% c("Saturday","Sunday"),"Weekend","Weekday"))

#calculate average of number of steps taken by time interval (accross all weekday days or weekend days )
dailyStepsIntervalMeanByDays <- aggregate(ActivityFillNA$steps, by=list(ActivityFillNA$interval,ActivityFillNA$wdays ),FUN=mean)


#specifiy the column names
colnames(dailyStepsIntervalMeanByDays) <- c("Interval","WDays","Steps")
dailyStepsIntervalMeanByDays$Steps <- round(dailyStepsIntervalMeanByDays$Steps)

library(ggplot2)

#draw the plot
g <- ggplot(dailyStepsIntervalMeanByDays, aes(Interval, Steps)) + 
  geom_line() +
  facet_grid(. ~ WDays)
    labs(title="Time Interval Series Vs Steps") 

print(g)
```
